---
title: "The two most important functions for data visualisation in R"
output: html_document
date: "2018-05-13T21:48:51-07:00"
category: 'R'
author: 'Mohammad Alfaifi'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
 
```


In my opinion, the two most important functions for data visualisation are *Par* and *Pairs* functions.

## Par Function
The first function is *par* function with *mfrow* parameter. The par(mfrow=c(rows,columns)) allows us to plot multiple plots on the same page or window. 

Here is an example with IRIS dataset. First, loading and inspecting the dataset 

```{r}
#import libraries

library(datasets)
#loading iris dataset
data(iris)
#The first six rows with the four columns
head(iris)

#plotting the different types of Species
#checking the class of the column
class(iris$Species)

#since it is a factor, its levels can be used to show the unique Species
levels(iris$Species) # there are three types: setosa, versicolor and virginica

```
Next, plotting the scatter plot of the sepal length and width using GGplot2.

```{r}
#import libraries

library(ggplot2)

# plotting sepal length and width as scatter plot

s<-ggplot(iris,aes(x=Sepal.Length, y=Sepal.Width))+geom_point()
s

```

Now,  I would like to plot the petal length and width
```{r}
p<-ggplot(iris,aes(x=Petal.Length,y=Petal.Width))+geom_point()
p
```

Now to compare both plots, I need  to go back and forth between the two or I can use the par function with mfrow parameter, which allows me to display both plots  side by side. Here is how it looks like

```{r eval=F}
par(mfrow=c(1,2)) # 1 is the number of rows while two is the number of columns
s # sepal length and width plot
p # petal length and width plot

```

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
multiplot <- function(..., plotlist=NULL, cols) {
    require(grid)

    # Make a list from the ... arguments and plotlist
    plots <- c(list(...), plotlist)

    numPlots = length(plots)

    # Make the panel
    plotCols = cols                          # Number of columns of plots
    plotRows = ceiling(numPlots/plotCols) # Number of rows needed, calculated from # of cols

    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(plotRows, plotCols)))
    vplayout <- function(x, y)
        viewport(layout.pos.row = x, layout.pos.col = y)

    # Make each plot, in the correct location
    for (i in 1:numPlots) {
        curRow = ceiling(i/plotCols)
        curCol = (i-1) %% plotCols + 1
        print(plots[[i]], vp = vplayout(curRow, curCol ))
    }

}

```

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
multiplot(s, p, cols=2)

```

We can change the number of rows and columns as we wish. Here is the same plots but using two rows and one column. 


```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
multiplot(s, p, cols=1)

```

# Pairs Function

The pairs function can be a very powerful tool for data exploration. Here is the iris dataset using pairs function.

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
pairs(iris)
```

The pairs function can be used to show different types of plots (using the pch argument) and/or show only a subset of the dataset. Here is an example of plotting the iris dataset with the sepal length column lower than 6. 
```{r}
pairs(iris, subset='Sepal.Length'<6 ,data=iris, pch=8)

```

Additionally, we can change the scale of the variables and plot them using only one line of code, which I found very useful when dealing with non-stationary data like the stock prices. 
```{r}


pairs(iris[1:4], log = "xy",pch=21, bg = c("red", "green", "blue")[unclass(iris$Species)]) 

```


This was it. If you have any suggestion or comments about this post, please let me know! 
