---
title: "Subseting Daily Time Series Data Using Data.Table Package"
output: html_document
date: "2018-05-24T21:48:51-07:00"
category: 'R'
author: 'Mohammad Alfaifi'

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
 
```


This blog post shows how to get the beginning (end) of the month (year) observations from daily time series data.  In my dissertation, I needed to calculate monthly returns and get the book-value and the market-value of the last day of June and December for different firms to construct Fama-French three-factor model. The easiest way I found to achieve this is through the code shown in this post. 

**The prerequisite for this short tutorial is to be familiar with data.table package syntax.**

The data loaded below is for the daily adjusted close price of S&P500 over the last year.
 

```{r message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}


sp500 <-readRDS('../../static/data/sp500.rds')
```

```{r message=FALSE, warning=FALSE, paged.print=FALSE}

head(sp500)
tail(sp500)
```


The first thing I want to do is to add two columns:  *year* and  *month*  using the zoo and lubridate packages so that I can use them later to subset the data frame. 

```{r}
#check the class of the data column

class(sp500$Date)
#changing its class to date
sp500$Date<-as.Date(sp500$Date)

#importing Zoo and lubridate
library(zoo)
library(lubridate)

sp500$year_month <- as.yearmon(sp500$Date) #add the year month column
sp500$year <- year(sp500$Date) # adding the year column

head(sp500)

```

## Last trading day of each month

Let's say that we need the last observation of each month. We may need this kind of data to calculate  the monthly returns of S&P500. We can achieve this using the groupby, and .SD in the j command of data table. The .SD returns a subset of x's data of each group. 

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
#importing data.table library

library(data.table)

sp500<-as.data.table(sp500) #chaning sp500

last_daymonth_value<- sp500[,.SD[.N],by=.(year_month)] #SD subsets the (.N) or last observation of each month

last_daymonth_value
```

## First trading day of each month

If we want the first value of each month we can change *'.N'* to *1* to get the desired results

```{r message=FALSE, warning=FALSE, paged.print=FALSE}

first_daymonth_value<- sp500[,.SD[1],by=.(year_month)] #SD subsets the (1) observation of each month

first_daymonth_value

```

## First observation of each year

The same thing can be done with the year column, below are the first observation of each year. 

```{r message=FALSE, warning=FALSE, paged.print=FALSE}

first_year_value<- sp500[,.SD[1],by=.(year)] #SD subsets the (1) observation of each year
first_year_value
```

## Last observation of each year

To get the last observation of each year, we need only to change *SD[1]* to *SD[.N]* as it is shown below

```{r message=FALSE, warning=FALSE, paged.print=FALSE}

last_year_value<- sp500[,.SD[.N],by=.(year)] #SD subsets the (N) or last observation of each year
last_year_value
```
## The second observation of each month


This technique does not only work with the first  or last trading day of the month(year).  We can get the code below to return the second value of each month as shown below

```{r message=FALSE, warning=FALSE, paged.print=FALSE}

last_year_value<- sp500[,.SD[2],by=.(year_month)] #SD subsets the (2) observation of each month

last_year_value
```




This was it. If you have any suggestion or comments, please let me know! 