---
title: "The two most important functions for data visualisation in R"
output: html_document
date: "2018-05-13T21:48:51-07:00"
category: 'R'
author: 'Mohammad Alfaifi'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
 
```


In my opinion, the two most important functions for data visualisation in R are *Par* and *Pairs* functions.

## Par Function
The first function is *par* function with *mfrow* parameter. The par(mfrow=c(rows,columns)) allows us to plot multiple plots on the same page or window. 

Below an example with IRIS dataset. First, loading and inspecting the dataset 

```{r}
#import libraries

library(datasets)
#loading iris dataset
data(iris)
#The first six rows with the four columns
head(iris)

#checking the class of the Species's column or feature
class(iris$Species)

#since its class is a factor, its levels can be used to show the unique values of the species column

levels(iris$Species) # there are three types: setosa, versicolor and virginica

```
Next, plotting the scatter plot of the sepal length and width using ggplot2 package. I am just randomly plotting different variables. The purpose will be clearer soon.

```{r}
#import libraries

library(ggplot2)

# plotting sepal length and width as scatter plot

s<-ggplot(iris,aes(x=Sepal.Length, y=Sepal.Width))+geom_point()
s

```

Now, plotting the petal length and width
```{r}
p<-ggplot(iris,aes(x=Petal.Length,y=Petal.Width))+geom_point()
p
```

To compare both plots, I need either to go back and forth between both plots or I can use the par function with mfrow parameter, which allows me to display both plots  side by side as it is shown below  

```{r eval=F}
par(mfrow=c(1,2)) # 1 is the number of rows while two is the number of columns
s # sepal length and width plot
p # petal length and width plot

```

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
multiplot <- function(..., plotlist=NULL, cols) {
    require(grid)

    # Make a list from the ... arguments and plotlist
    plots <- c(list(...), plotlist)

    numPlots = length(plots)

    # Make the panel
    plotCols = cols                          # Number of columns of plots
    plotRows = ceiling(numPlots/plotCols) # Number of rows needed, calculated from # of cols

    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(plotRows, plotCols)))
    vplayout <- function(x, y)
        viewport(layout.pos.row = x, layout.pos.col = y)

    # Make each plot, in the correct location
    for (i in 1:numPlots) {
        curRow = ceiling(i/plotCols)
        curCol = (i-1) %% plotCols + 1
        print(plots[[i]], vp = vplayout(curRow, curCol ))
    }

}

```

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
multiplot(s, p, cols=2)

```

We can change the number of rows and columns as we wish. Below is the same plots but using *two rows* and *one column*. 


```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
multiplot(s, p, cols=1)

```

# Pairs Function

The pairs function can be a very powerful tool for data exploration. Below is the iris dataset plotted using pairs function.

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
pairs(iris)
```

The pairs function can be used to show different types of plots (using the pch argument) and/or show only a subset of the dataset. Here is an example of plotting the iris dataset with the sepal length column lower than 6. 
```{r}
pairs(iris, subset='Sepal.Length'<6 ,data=iris, pch=8)

```

Additionally, we can change the scale of the variables and plot them using only one line of code, which I find it very useful when dealing with non-stationary data like the financial data.
```{r}


pairs(iris[1:4], log = "xy",pch=21, bg = c("red", "green", "blue")[unclass(iris$Species)]) 

```

The Sepal.Length is the x-axis for each plot that it below the text plot 'Sepal.Length' while it is the y-axis for each plot that it is on the right of the text plot. 



This was it.  If you have any suggestion or comments about this post, please let me know! 
